### 2021-10-11 papers/thesis/resubmitted/ch4-spherical-caps/spherical-caps.tex
Fixed commas, fullstops etc. that follow equations to make consistent in CH4.

Our aim in this chapter is to develop a sparse spectral method for solving linear partial differential equations on certain subsets of the sphere -- specifically spherical caps. More precisely, we consider the solution of partial differential equations on the \textit{spherical cap} $\Omega$ defined by
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{\alpha < z < \beta, \: x^2 + y^2 + z^2 = 1}
}
where $\alpha \in (-1,1)$ and $\beta := 1$. Simply put, the region of the surface of the $2$-sphere where the $z$-coordinate range is limited to the interval $[\alpha,1]$ is what we refer to as the spherical cap.

@@

On the spherical cap, the family of weights we will consider are of the form
\bseqn{
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \Omega$ when $a > 0$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k-i} \: y^i \: z^{n-k} + (\hbox{lower order terms})
}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - i} \: y^{i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs that we will see in \bsrefdef{def:sc:OPconstruction}.

Just as for the spherical harmonics framework for the whole sphere in \bsrefchapter{CHAPTER:sphericalharmonics}, we consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator). Recall that we write the spherical coordinates as
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi
}
where $ \rho(z) := \sqrt{1-z^2}$, and that the Laplace--Beltrami operator is then
\bseqn{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for $\xvec := (x,y,z) \in \R^3$. We do so by considering the component operators $\rho {\partial \over \partial \varphi}$ and ${\partial \over \partial \theta}$ applied to OPs with a specific choices of weight so that their discretisation is sparse, see \bsreftheorem{theorem:sc:sparsityofdifferentialoperators}. Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$, \`a la the ultraspherical spectral method we investigated in \bsrefchapter{CHAPTER:diskslice} for disk-slices and trapeziums (see also \cite{snowball2019sparse}), as well as for intervals \cite{olver2013fast}, triangles \cite{olver2019triangle}, and the related work on sparse discretisations involving Jacobi polynomials on disks \cite{vasil2016tensor} and spheres \cite{vasil2019tensor,lecoanet2019tensor}. Just as we proceeded in \bsrefchapter{CHAPTER:diskslice} for the disk-slice case in 2D (see also \cite{snowball2019sparse}), we will use an integration-by-parts argument to deduce the sparsity structure.

@@

The spherical cap can be thought of as a higher dimensional version of the circle arc (a one-dimensional \enquote{surface} in two-dimensional space). Recently, a procedure for defining OPs in two variables on quadratic curves, such as ellipses and parabolas, has been presented \cite{olver2021orthogonal}, where the unit circle is one of the five classes of curve that can be generalised to give any curve in the plane $\R^2$. One can define OPs on the circle in terms of Fourier series, which we write here as orthogonal polynomials in $x$ and $y$.
\begin{definition}[\cite{olver2021orthogonal}]\label{def:sc:Ydefinition}
	Define the unit circle $\omega := \bsset{\vec{x} = (x,y) \in \R^2}{x^2 + y^2 = 1}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\bseqn{
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta) \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta)
}
for $k = 1,2,3,\dots$ where $\ch_{0} := {\sqrt{2} \over 2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\bseqn{
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta.
}
Note that $\ch_0$ is defined so as to ensure orthonormality. 
\end{definition}

However, for an arc of the unit circle $\bsset{\vec{x} = (x,y) \in \R^2}{x \ge h, \: x^2 + y^2 = 1}$ for some $h \in (-1, 1)$ (note we can orientate the arc as we please without loss of generality), we would need to make a modification so that the polynomials are orthonormal with respect to a different inner product accounting for the truncated domain of the arc:
\bseqn{
	\ip< p, \: q >_{\ch^h} &:= c^h \: \int_{-\arccos{h}}^{\arccos{h}} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta \\
	&= c^h \: \int_h^1 \Big\{ p(x,\sqrt{1-x^2}) \; q(x,\sqrt{1-x^2}) \\
	&\quadfour \quadtwo + p(x,-\sqrt{1-x^2}) \; q(x,-\sqrt{1-x^2}) \Big\} {\D x \over \sqrt{1-x^2}}
}
for some normalising constant $c^h$. In the same vein, we can choose to construct the arc OPs in the same way as the whole circle OPs -- that is, define $\{T_k^h\}$, $\{U_k^h\}$ as two sets of univariate orthogonal polynomials on the interval $[h,1]$, orthogonal with respect to the weight functions $(1-x^2)^{-\half}$ and $(1-x^2)^{\half}$ respectively, and then define
\bseqn{
	\ch^h_{0,0}(\xvec) \equiv \ch^h_{0,0}(x, y) &:= \ch^h_{0} =: \ch^h_{0,0}(\theta) \\
	\ch^h_{k,0}(\xvec) \equiv \ch^h_{k,0}(x, y) &:= T^h_k(x) =: \ch^h_{k,0}(\theta) \\
	\ch^h_{k,1}(\xvec) \equiv \ch^h_{k,1}(x, y) &:= y \: U^h_{k-1}(x) =: \ch^h_{k,1}(\theta)
}
where $Y_0^h$ is chosen so that $\ip< Y_0^h, \: Y_0^h >_{\ch^h} = 1$.

@@

Using this approach, we can construct the 3D orthogonal polynomials on the spherical cap or band $\Omega$ from 1D orthogonal polynomials on the interval $[\alpha,\beta]$, and from Fourier series written as orthogonal polynomials in $x$ and $y$. Recall that the spherical cap is defined by
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{\alpha < z < \beta, \: x^2 + y^2 + z^2 = 1}.
}
\begin{proposition}[\cite{olver2020orthogonal}]\label{prop:sc:OPconstruction}
	Let $w : (\alpha,\beta) \to \R$ be a weight function. For $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$ given by
\bseqn{
	\scopnki(x,y,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr({x \over \rho(z)}, {y \over \rho(z)})
}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\bseqn{
	&\ip< p, \: q > := \int_\Omega p(x,y,z) \: q(x,y,z) \: w(z) \: \D A \\
	&= \int_0^{\cos^{-1}(\alpha)} \int_0^{2\pi} \Big\{ p\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: q\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \\
	&\quadeight \quad \cdot w(\cosphi) \: \sinphi \Big\} \: \D \theta \: \D \varphi \\
	&= \int_\alpha^1 \int_0^{2\pi} p\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: q\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: w(z) \: \D \theta \: \D z
}
on $\Omega$, where $\D A = \sinphi \: \D \theta \: \D \varphi$ is the uniform spherical measure on $\Omega$. 
\end{proposition}

For the spherical cap, we can use \bsrefprop{prop:sc:OPconstruction} to create our one-parameter family of OPs. The univariate OPs that we will choose for the $r_{n,k}$ polynomials above will be the non-classical $\genjac^{(a,b)}$ OPs that we defined in \bsrefdef{def:ds:OPconstruction}. Since there is only one boundary for the spherical cap, we will only need to use a two-parameter version\footnote{For a spherical band with two boundaries, we would need the three-parameter version}). For reference, the family of orthonormal polynomials on $[\alpha,\beta]$ denoted $\{\genjac_n^{(a,b)}\}$ are defined such that they are orthonormal with respect to the inner product
\bseqnnumber{
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:sc:ipgenjac}
}
where
\bseqn{
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{a} \: \rho(x)^{b}
} is the weight function and 
\bseqnnumber{
	\normgenjac^{(a,b)} := \int_\alpha^1 \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:sc:ipnormalisation}
}
is a normalising constant.

@@

Recall that we can express the three-term recurrences associated with $\genjac_n^{(a,b)}$ as
\bseqnnumber{
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x) \label{eqn:sc:Rrecurrence}
}
where the coefficients are calculable (see \bsrefchapter{CHAPTER:diskslice}). We can use \bsrefeqn{eqn:sc:Rrecurrence} to determine the 3D recurrences for $\scopnkia(x,y,z)$. Importantly, we can deduce sparsity in the recurrence relationships.  We first require the following lemma.

@@

\begin{proof}
Each follows from the definitions of $\ch_{k,i}$ and $\ch_0$, as well as the trigonometric relationships:
\bseqn{
	2 \cos k \theta \cos\theta &= \cos (k-1)\theta + \cos(k+1)\theta \\
	2 \sin k \theta \cos\theta &= \sin (k-1)\theta + \sin(k+1)\theta \\
	2 \cos k \theta \sin\theta &= - \sin (k-1)\theta + \sin(k+1)\theta \\
	2 \sin k \theta \sin\theta &= \cos (k-1)\theta - \cos(k+1)\theta.
}
\end{proof}

\begin{lemma}\label{lemma:sc:Qrecurrence} 
Define
\bseqnnumber{
	\eta_{k} :=
		\begin{cases}
			0 &\text{if } k < 0 \\
			\ch_0 &\text{if } k = 0 \\
			\half &\text{otherwise}
		\end{cases}
}
$\scopnkia(x,y,z)$ satisfy the following recurrences:
\bseqn{
	x \: \scopnkia(x,y,z) &= \alphaa_{n,k,1} \:  \scopa_{n-1, k-1, i}(x, y, z) + \alphaa_{n,k,2} \:  \scopa_{n-1, k+1, i}(x, y, z) \nonumber \\
		& \quad + \alphaa_{n,k,3} \:  \scopa_{n, k-1, i}(x, y, z) + \alphaa_{n,k,4} \:  \scopa_{n, k+1, i}(x, y, z) \nonumber \\
		& \quad + \alphaa_{n,k,5} \:  \scopa_{n+1, k-1, i}(x, y, z) + \alphaa_{n,k,6} \:  \scopa_{n+1, k+1, i}(x, y, z), \\ \\
	y \: \scopnkia(x,y,z) &= \betaa_{n,k,i,1} \:  \scopa_{n-1, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,2} \:  \scopa_{n-1, k+1, |i-1|}(x, y, z) \nonumber \\
		& \quad + \betaa_{n,k,i,3} \:  \scopa_{n, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,4} \:  \scopa_{n, k+1, |i-1|}(x, y, z) \nonumber \\
		& \quad + \betaa_{n,k,i,5} \:  \scopa_{n+1, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,6} \:  \scopa_{n+1, k+1, |i-1|}(x, y, z), \\ \\
	z \: \scopnkia(x,y,z) &= \gammaa_{n,k,1} \: \scopa_{n-1, k, i}(x, y, z) + \gammaa_{n,k,2} \: \scopa_{n, k, i}(x, y, z) \nonumber \\
		& \quad + \gammaa_{n,k,3} \: \scopa_{n+1, k, i}(x, y, z),
}
for $(x,y,z) \in \Omega$, where
\bseqn{
	\alphaa_{n,k,1} &:= \eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjacnmk^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,2} &:= \eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjac_{n-k-2}^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}}, \\
	\alphaa_{n,k,3} &:=\eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjac_{n-k+1}^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,4} &:= \eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjac_{n-k-1}^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}}, \\
	\alphaa_{n,k,5} &:= \eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjac_{n-k+2}^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,6} &:=\eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjacnmk^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}},
}
\bseqn{
	\betaa_{n,k,i,j} &:= 
		\begin{cases}
			- \alphaa_{n,k,j} \quad &\text{if }(i = 0 \text{ and } j \text{ is odd}) \text{ or }(i = 1 \text{ and } j \text{ is even}) \\
			\alphaa_{n,k,j} \quad &\text{otherwise}
		\end{cases}, \\	
	\gammaa_{n,k,1} &:= \beta_{n-k-1}^{(a, 2k)}, \qquad \gammaa_{n,k,2} := \alpha_{n-k}^{(a, 2k)}, \qquad \gammaa_{n,k,3} := \beta_{n-k}^{(a, 2k)}.
}
\end{lemma}

@@

For each $n = 0,1,2\dots$ let $\Dnt$ be any matrix that is a left inverse of $A_n$, i.e. such that $\Dnt A_n = I_{2n+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dnt$'s, we obtain a lower triangular system \cite[p.78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence for $n = 0,1,2,\dots$:
\bseqn{
	\begin{cases}
		\bigscopta_{-1}(x,y,z) := 0 \\
		\bigscopta_{0}(x,y,z) := \scopa_0 \\
		\bigscopta_{n+1}(x,y,z) = -\Dnt (B_n-G_n(x,y,z)) \bigscopta_n(x,y,z) - \Dnt C_n  \, \bigscopta_{n-1}(x,y,z)
	\end{cases}.
}

>>>>

Our aim in this chapter is to develop a sparse spectral method for solving linear partial differential equations on certain subsets of the sphere -- specifically spherical caps. More precisely, we consider the solution of partial differential equations on the \textit{spherical cap} $\Omega$ defined by
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{\alpha < z < \beta, \: x^2 + y^2 + z^2 = 1},
}
where $\alpha \in (-1,1)$ and $\beta := 1$. Simply put, the region of the surface of the $2$-sphere where the $z$-coordinate range is limited to the interval $[\alpha,1]$ is what we refer to as the spherical cap.

@@

On the spherical cap, the family of weights we will consider are of the form
\bseqn{
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \Omega$ when $a > 0$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\bseqn{
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k-i} \: y^i \: z^{n-k} + (\hbox{lower order terms}),
}
where $C_{n,k,i} \neq 0$ and \enquote{lower order terms} includes degree $n$ polynomials of the form $x^{j - i} \: y^{i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs that we will see in \bsrefdef{def:sc:OPconstruction}.

Just as for the spherical harmonics framework for the whole sphere in \bsrefchapter{CHAPTER:sphericalharmonics}, we consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator). Recall that we write the spherical coordinates as
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta, \\
	y &= \sinphi \sintheta = \rho(z) \sintheta, \\
	z &= \cosphi,
}
where $ \rho(z) := \sqrt{1-z^2}$, and that the Laplace--Beltrami operator is then
\bseqn{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo,
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for $\xvec := (x,y,z) \in \R^3$. We do so by considering the component operators $\rho {\partial \over \partial \varphi}$ and ${\partial \over \partial \theta}$ applied to OPs with a specific choices of weight so that their discretisation is sparse, see \bsreftheorem{theorem:sc:sparsityofdifferentialoperators}. Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$, \`a la the ultraspherical spectral method we investigated in \bsrefchapter{CHAPTER:diskslice} for disk-slices and trapeziums (see also \cite{snowball2019sparse}), as well as for intervals \cite{olver2013fast}, triangles \cite{olver2019triangle}, and the related work on sparse discretisations involving Jacobi polynomials on disks \cite{vasil2016tensor} and spheres \cite{vasil2019tensor,lecoanet2019tensor}. Just as we proceeded in \bsrefchapter{CHAPTER:diskslice} for the disk-slice case in 2D (see also \cite{snowball2019sparse}), we will use an integration-by-parts argument to deduce the sparsity structure.

@@

The spherical cap can be thought of as a higher dimensional version of the circle arc (a one-dimensional \enquote{surface} in two-dimensional space). Recently, a procedure for defining OPs in two variables on quadratic curves, such as ellipses and parabolas, has been presented \cite{olver2021orthogonal}, where the unit circle is one of the five classes of curve that can be generalised to give any curve in the plane $\R^2$. One can define OPs on the circle in terms of Fourier series, which we write here as orthogonal polynomials in $x$ and $y$.
\begin{definition}[\cite{olver2021orthogonal}]\label{def:sc:Ydefinition}
	Define the unit circle $\omega := \bsset{\vec{x} = (x,y) \in \R^2}{x^2 + y^2 = 1}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\bseqn{
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta), \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta), \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta),
}
for $k = 1,2,3,\dots$ where $\ch_{0} := {\sqrt{2} \over 2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\bseqn{
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta.
}
Note that $\ch_0$ is defined so as to ensure orthonormality. 
\end{definition}

However, for an arc of the unit circle $\bsset{\vec{x} = (x,y) \in \R^2}{x \ge h, \: x^2 + y^2 = 1}$ for some $h \in (-1, 1)$ (note we can orientate the arc as we please without loss of generality), we would need to make a modification so that the polynomials are orthonormal with respect to a different inner product accounting for the truncated domain of the arc:
\bseqn{
	\ip< p, \: q >_{\ch^h} &:= c^h \: \int_{-\arccos{h}}^{\arccos{h}} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta \\
	&= c^h \: \int_h^1 \Big\{ p(x,\sqrt{1-x^2}) \; q(x,\sqrt{1-x^2}) \\
	&\quadfour \quadtwo + p(x,-\sqrt{1-x^2}) \; q(x,-\sqrt{1-x^2}) \Big\} {\D x \over \sqrt{1-x^2}},
}
for some normalising constant $c^h$. In the same vein, we can choose to construct the arc OPs in the same way as the whole circle OPs -- that is, define $\{T_k^h\}$, $\{U_k^h\}$ as two sets of univariate orthogonal polynomials on the interval $[h,1]$, orthogonal with respect to the weight functions $(1-x^2)^{-\half}$ and $(1-x^2)^{\half}$ respectively, and then define
\bseqn{
	\ch^h_{0,0}(\xvec) \equiv \ch^h_{0,0}(x, y) &:= \ch^h_{0} =: \ch^h_{0,0}(\theta), \\
	\ch^h_{k,0}(\xvec) \equiv \ch^h_{k,0}(x, y) &:= T^h_k(x) =: \ch^h_{k,0}(\theta), \\
	\ch^h_{k,1}(\xvec) \equiv \ch^h_{k,1}(x, y) &:= y \: U^h_{k-1}(x) =: \ch^h_{k,1}(\theta),
}
where $Y_0^h$ is chosen so that $\ip< Y_0^h, \: Y_0^h >_{\ch^h} = 1$.

@@

Using this approach, we can construct the 3D orthogonal polynomials on the spherical cap or band $\Omega$ from 1D orthogonal polynomials on the interval $[\alpha,\beta]$, and from Fourier series written as orthogonal polynomials in $x$ and $y$. Recall that the spherical cap is defined by
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{\alpha < z < \beta, \: x^2 + y^2 + z^2 = 1}.
}
\begin{proposition}[\cite{olver2020orthogonal}]\label{prop:sc:OPconstruction}
	Let $w : (\alpha,\beta) \to \R$ be a weight function. For $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$ given by
\bseqn{
	\scopnki(x,y,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr({x \over \rho(z)}, {y \over \rho(z)}),
}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\bseqn{
	&\ip< p, \: q > := \int_\Omega p(x,y,z) \: q(x,y,z) \: w(z) \: \D A \\
	&= \int_0^{\cos^{-1}(\alpha)} \int_0^{2\pi} \Big\{ p\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: q\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \\
	&\quadeight \quad \cdot w(\cosphi) \: \sinphi \Big\} \: \D \theta \: \D \varphi \\
	&= \int_\alpha^1 \int_0^{2\pi} p\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: q\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: w(z) \: \D \theta \: \D z,
}
on $\Omega$, where $\D A = \sinphi \: \D \theta \: \D \varphi$ is the uniform spherical measure on $\Omega$. 
\end{proposition}

For the spherical cap, we can use \bsrefprop{prop:sc:OPconstruction} to create our one-parameter family of OPs. The univariate OPs that we will choose for the $r_{n,k}$ polynomials above will be the non-classical $\genjac^{(a,b)}$ OPs that we defined in \bsrefdef{def:ds:OPconstruction}. Since there is only one boundary for the spherical cap, we will only need to use a two-parameter version\footnote{For a spherical band with two boundaries, we would need the three-parameter version}). For reference, the family of orthonormal polynomials on $[\alpha,\beta]$ denoted $\{\genjac_n^{(a,b)}\}$ are defined such that they are orthonormal with respect to the inner product
\bseqnnumber{
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \label{eqn:sc:ipgenjac}
}
where
\bseqn{
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{a} \: \rho(x)^{b},
} 
is the weight function and 
\bseqnnumber{
	\normgenjac^{(a,b)} := \int_\alpha^1 \: \genjacw^{(a,b)}(x) \: \D x, \label{eqn:sc:ipnormalisation}
}
is a normalising constant.

@@

Recall that we can express the three-term recurrences associated with $\genjac_n^{(a,b)}$ as
\bseqnnumber{
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x), \label{eqn:sc:Rrecurrence}
}
where the coefficients are calculable (see \bsrefchapter{CHAPTER:diskslice}). We can use \bsrefeqn{eqn:sc:Rrecurrence} to determine the 3D recurrences for $\scopnkia(x,y,z)$. Importantly, we can deduce sparsity in the recurrence relationships.  We first require the following lemma.

@@

\begin{proof}
Each follows from the definitions of $\ch_{k,i}$ and $\ch_0$, as well as the trigonometric relationships:
\bseqn{
	2 \cos k \theta \cos\theta &= \cos (k-1)\theta + \cos(k+1)\theta, \\
	2 \sin k \theta \cos\theta &= \sin (k-1)\theta + \sin(k+1)\theta, \\
	2 \cos k \theta \sin\theta &= - \sin (k-1)\theta + \sin(k+1)\theta, \\
	2 \sin k \theta \sin\theta &= \cos (k-1)\theta - \cos(k+1)\theta.
}
\end{proof}

\begin{lemma}\label{lemma:sc:Qrecurrence} 
Define
\bseqnnumber{
	\eta_{k} :=
		\begin{cases}
			0 &\text{if } k < 0 \\
			\ch_0 &\text{if } k = 0 \\
			\half &\text{otherwise}.
		\end{cases}
}
$\scopnkia(x,y,z)$ satisfy the following recurrences:
\bseqn{
	x \: \scopnkia(x,y,z) &= \alphaa_{n,k,1} \:  \scopa_{n-1, k-1, i}(x, y, z) + \alphaa_{n,k,2} \:  \scopa_{n-1, k+1, i}(x, y, z) \nonumber \\
		& \quad + \alphaa_{n,k,3} \:  \scopa_{n, k-1, i}(x, y, z) + \alphaa_{n,k,4} \:  \scopa_{n, k+1, i}(x, y, z) \nonumber \\
		& \quad + \alphaa_{n,k,5} \:  \scopa_{n+1, k-1, i}(x, y, z) + \alphaa_{n,k,6} \:  \scopa_{n+1, k+1, i}(x, y, z), \\ \\
	y \: \scopnkia(x,y,z) &= \betaa_{n,k,i,1} \:  \scopa_{n-1, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,2} \:  \scopa_{n-1, k+1, |i-1|}(x, y, z) \nonumber \\
		& \quad + \betaa_{n,k,i,3} \:  \scopa_{n, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,4} \:  \scopa_{n, k+1, |i-1|}(x, y, z) \nonumber \\
		& \quad + \betaa_{n,k,i,5} \:  \scopa_{n+1, k-1, |i-1|}(x, y, z) + \betaa_{n,k,i,6} \:  \scopa_{n+1, k+1, |i-1|}(x, y, z), \\ \\
	z \: \scopnkia(x,y,z) &= \gammaa_{n,k,1} \: \scopa_{n-1, k, i}(x, y, z) + \gammaa_{n,k,2} \: \scopa_{n, k, i}(x, y, z) \nonumber \\
		& \quad + \gammaa_{n,k,3} \: \scopa_{n+1, k, i}(x, y, z),
}
for $(x,y,z) \in \Omega$, where
\bseqn{
	\alphaa_{n,k,1} &:= \eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjacnmk^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,2} &:= \eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjac_{n-k-2}^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}}, \\
	\alphaa_{n,k,3} &:=\eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjac_{n-k+1}^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,4} &:= \eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjac_{n-k-1}^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}}, \\
	\alphaa_{n,k,5} &:= \eta_{k-1} \: \ip<\genjacnmk^{(a, 2k)}, \: \genjac_{n-k+2}^{(a, 2(k-1))}>_{\genjacw^{(a, 2k)}}, \\
	\alphaa_{n,k,6} &:=\eta_{k} \: \ip<\genjacnmk^{(a, 2k)}, \genjacnmk^{(a, 2(k+1))}>_{\genjacw^{(a, 2(k+1))}},
}
\bseqn{
	\betaa_{n,k,i,j} &:= 
		\begin{cases}
			- \alphaa_{n,k,j} \quad &\text{if }(i = 0 \text{ and } j \text{ is odd}) \text{ or }(i = 1 \text{ and } j \text{ is even}) \\
			\alphaa_{n,k,j} \quad &\text{otherwise},
		\end{cases} \\	
	\gammaa_{n,k,1} &:= \beta_{n-k-1}^{(a, 2k)}, \qquad \gammaa_{n,k,2} := \alpha_{n-k}^{(a, 2k)}, \qquad \gammaa_{n,k,3} := \beta_{n-k}^{(a, 2k)}.
}
\end{lemma}

@@

For each $n = 0,1,2\dots$ let $\Dnt$ be any matrix that is a left inverse of $A_n$, i.e. such that $\Dnt A_n = I_{2n+3}$. Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the $\Dnt$'s, we obtain a lower triangular system \cite[p.78]{dunkl2014orthogonal}, which can be expanded to obtain the recurrence for $n = 0,1,2,\dots$:
\bseqn{
	\begin{cases}
		\bigscopta_{-1}(x,y,z) := 0, \\
		\bigscopta_{0}(x,y,z) := \scopa_0, \\
		\bigscopta_{n+1}(x,y,z) = -\Dnt (B_n-G_n(x,y,z)) \bigscopta_n(x,y,z) - \Dnt C_n  \, \bigscopta_{n-1}(x,y,z).
	\end{cases}
}
